---
title: "Getting Started with memtoc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with memtoc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction
 
memtoc provides simple start/stop memory tracking for R, inspired by the
[tictoc](https://github.com/jabiru/tictoc) package for timing. Wrap any code
block with `tic_mem()` and `toc_mem()` to measure RAM usage.

```{r setup}
library(memtoc)
```

## Basic Usage

The simplest use case is tracking memory for a single operation:

```{r basic}
tic_mem("load data")
data <- read.csv("large_file.csv")
toc_mem()
#> ✔ load data: 142.3 MB peak | 142.3 MB current | 1.24 sec | 2 samples
```

The output shows:

- **Peak memory**: Maximum RSS (Resident Set Size) during the operation
- **Current memory**: RSS at the end of the operation
- **Elapsed time**: Wall-clock time
- **Samples**: Number of memory measurements taken

## Background Polling

By default, memtoc spawns a background process that continuously samples memory.
This captures the true peak even for short-lived allocations:
 
```{r polling}
tic_mem("matrix operations", interval = 0.5)  # Sample every 0.5 seconds

# Create a large temporary matrix
x <- matrix(rnorm(1e8), ncol = 1000)  # ~800 MB
y <- colMeans(x)                        # x can be garbage collected
rm(x)
gc()

result <- toc_mem()
#> ✔ matrix operations: 812.4 MB peak | 45.2 MB current | 3.21 sec | 7 samples
```

Without background polling, you would only see the final memory (45.2 MB),
missing the 800 MB peak. Access the full trajectory with `result$trajectory`.

For very quick operations, disable polling to avoid startup overhead:

```{r snapshot}
tic_mem("quick op", interval = NULL)  # Snapshot mode
y <- 1:100
toc_mem()
```

## Nested Tracking

Track an entire pipeline while also measuring individual steps:

```{r nested}
tic_mem("full pipeline")

  tic_mem("step 1: load")
  data <- read.csv("data.csv")
  toc_mem()
  #> ✔ step 1: load: 50.2 MB peak | 50.2 MB current | 1.2 sec

  tic_mem("step 2: transform")
  features <- transform(data)
  toc_mem()
  #> ✔ step 2: transform: 125.8 MB peak | 98.3 MB current | 2.4 sec

  tic_mem("step 3: model")
  model <- train(features)
  toc_mem()
  #> ✔ step 3: model: 512.1 MB peak | 201.5 MB current | 45.2 sec

toc_mem()
#> ✔ full pipeline: 512.1 MB peak | 201.5 MB current | 48.8 sec
```

## Logging Results

Collect results for later analysis:

```{r logging}
mem_clearlog()

for (i in 1:10) {
  tic_mem(paste("iteration", i))
  # ... do work ...
  toc_mem(log = TRUE, quiet = TRUE)
}

# Get all results as a data frame
results <- mem_log()
summary(results$mem_peak)
```

## Parallel Worker Monitoring

When using the `future` package for parallel processing, memtoc can monitor
memory across all workers:

```{r parallel}
library(future)
library(future.apply)

# Set up parallel workers
plan(multisession, workers = 4)

# Check that workers are detected
mem_parallel_info()
#> ── Parallel Backend Info
#> • Main process PID: 12345
#> • Current plan: multisession
#> • Workers configured: 4

# Monitor parallel job
tic_mem("parallel computation", workers = "auto")
result <- future_lapply(1:100, function(i) {
  x <- rnorm(1e6)
  mean(x)
}, future.seed = TRUE)
mem_result <- toc_mem()
#> ✔ parallel computation: 1.2 GB peak | 245 MB current | 5.4 sec | 4 workers

# View per-worker breakdown
mem_result$worker_stats

# Clean up
plan(sequential)
```

Worker options:

- `workers = "auto"`: Auto-detect future workers (default)
- `workers = "none"`: Only monitor main process
- `workers = "children"`: Monitor main process and child processes
- `workers = c(pid1, pid2)`: Explicit PID list

## System Memory Warnings

memtoc warns you when system RAM is running low:

```{r warnings}
tic_mem("memory intensive")
# ... allocate lots of memory ...
toc_mem()
#> ✔ memory intensive: 12.4 GB peak | 11.2 GB current | 45.2 sec
#> ⚠ System RAM high: 87.3% used
```

Warnings appear at 80% usage; critical alerts at 95%.

## Crash Recovery

If R crashes during a long computation, recover the collected samples:

```{r recovery}
# After restarting R
mem_recover()
#> ℹ Found 1 recovery file:
#>   • PID 12345: 15.2 KB, 152 samples

# Recover the data
recovered <- mem_recover(pid = 12345)
head(recovered)
```

## Diagnostics

If background polling isn't working, run diagnostics:

```{r diagnostics}
mem_capabilities()
#>     memory_queries background_polling 
#>               TRUE               TRUE

# Detailed troubleshooting
mem_diagnose()
```

## Tips

1. **Use labels**: Always pass a message to `tic_mem()` for easier tracking
2. **Adjust interval**: Use shorter intervals (0.1-0.5s) for fast operations,
   longer intervals (1-5s) for long-running jobs
3. **Disable polling for quick ops**: Use `interval = NULL` for sub-second operations
4. **Monitor workers**: Set `workers = "auto"` when using `future` for parallelism
5. **Log results**: Use `log = TRUE` when running benchmarks or comparisons
